---
title: "Setter Decision and Hit Result in Volleyball Analysis"
author: "Julia Fung"
format: pdf
editor: visual
---

#Set up

Packages
```{r}
#| output: false
library(tidyverse)
library(rsample)
library(readxl)
library(PerformanceAnalytics)
library(nnet)
library(sjPlot)
library(effects)
library(randomForest)
library(gridExtra)
library(xtable)
```

Data
```{r}
#| output: false
setters <- read_excel("C:/Users/jfung/Files/Datasets/setters.xlsx")
```

Set Variable Types
```{r}
setters <- setters |>
  mutate(kill = as.logical(kill),
         rotation = as.factor(rotation),
         receiver_loc = as.factor(receiver_loc),
         pass_loc = as.factor(pass_loc),
         pass_quality = as.factor(pass_quality),
         set_loc = as.factor(set_loc),
         set_receiver = as.factor(set_receiver),
         team = as.factor(team),
         level = as.factor(level),
         game = as.factor(game),
         game_sex = as.factor(game_sex))
```

Split
```{r}
#| output: false
set.seed(3905)
df_split <- initial_split(setters,
                          prop = 0.75)
train <- training(df_split)
test <- testing(df_split)
```


# Data Exploration and Cleaning

## Numerical Predictors
```{r}
num_vars <- train[, c(2:7, 14)]
```

Correlation Matrix
```{r}
png(
file.path("_assets",
"num_var_corplot.png"),
width=1080,
height=800,
pointsize=30 )

par(mar=c(4, 4, 4, 1))
chart.Correlation(num_vars, 
                  histogram = TRUE)
mtext("Numerical Variable Relationships and Distributions", side = 3, line = 3)
dev.off()
```
Point variables are highly correlated, so only retain total_points and team1_diff

### Updated Data Set(train_clean)
```{r}
train_clean <- train |>
  select(-c(points1, points2, team2_diff))
```

## Categorical Predictors
```{r}
cat_vars <- train_clean[, -c(1:4, 10:13)]
```

Distributions Plot
```{r}
plots <- list()
for(var in names(cat_vars)){
  freqs <- data.frame(table(train_clean[var]))
  freqs$prop <- freqs$Freq/sum(freqs$Freq)
  plots[[var]] <- freqs |>
    ggplot(aes(x = !!sym(var), y = prop)) +
    geom_bar(stat = "identity") +
    ggtitle(paste(var, "Distribution")) + 
    theme_minimal()
}

png(
file.path("_assets",
"cat_var_dists.png"),
width=1080,
height=800,
pointsize=30 )

grid.arrange(grobs = plots,
             nrow = 3,
             ncol = 3)

dev.off()
```
Counts for some pass locations and qualities are low, so I will create a binary in-system variable.  Also 6 rotation to back and front binary

### Updated Data set
```{r}
train_clean <- train_clean |>
  mutate(insys = ifelse(pass_loc %in% c(2,3,4,7) & pass_quality %in% c(4,5),
                        TRUE,
                        FALSE),
         back_setter = ifelse(rotation %in% c(1, 2, 3),
                              TRUE,
                              FALSE)) |>
  select(-c(pass_loc, pass_quality, receiver_loc, rotation))
```

## Response Variables

### Set Location
```{r}
png(
file.path("_assets",
"set_loc_dist.png"),
width=1080,
height=800,
pointsize=30 )

par(mar=c(4, 4, 4, 1))
data.frame(table(train_clean$set_loc)) |>
    ggplot(aes(x = Var1, y = Freq)) +
    geom_bar(stat = "identity") +
    labs(title = "Set Location Distribution",
         x = "Set Location",
         y = "Frequency") + 
    theme_minimal()
dev.off()
```
Group infrequent locations

### Updated Set Locations
```{r}
train_clean <- train_clean |>
  mutate(set_loc = case_when(
    set_loc %in% c("back1", "1", "2", "back2", "3", "dump") ~ "midCourt",
    set_loc == "back3" ~ "5",
    set_loc %in% c("A", "B", "C", "D", "pipe") ~ "backRow",
    TRUE ~ set_loc
  )) |>
  mutate(set_loc = as.factor(set_loc))
# make a 4 the reference category(default set)
train_clean$set_loc <- relevel(train_clean$set_loc, ref = "4")
```

Distribution
```{r}
png(
file.path("_assets",
"set_loc2_dist.png"),
width=1080,
height=800,
pointsize=30 )

par(mar=c(4, 4, 4, 1))
data.frame(table(train_clean$set_loc)) |>
    ggplot(aes(x = Var1, y = Freq)) +
    geom_bar(stat = "identity") +
    labs(title = "Grouped Set Location Distribution",
         x = "Set Location",
         y = "Frequency") + 
    theme_minimal()
dev.off()
```

### Set Receiver
```{r}
png(
file.path("_assets",
"set_rec_dist.png"),
width=1080,
height=800,
pointsize=30 )

par(mar=c(4, 4, 4, 1))
data.frame(table(train_clean$set_receiver)) |>
    ggplot(aes(x = Var1, y = Freq)) +
    geom_bar(stat = "identity") +
    labs(title = "Set Receiver Distribution",
         x = "Set Receiver",
         y = "Frequency") + 
    theme_minimal()
dev.off()
```

### Updated Set Receivers
```{r}
train_clean <- train_clean |>
  mutate(set_receiver = case_when(
    set_receiver %in% c("ds", "libero", "setter1", "setter2", "opposite2", "outside3") ~ "other",
    set_receiver %in% c("middle1", "middle2", "middle3") ~ "middle",
    set_receiver == "outside" ~ "outside2",
    TRUE ~ set_receiver
  ))
train_clean$set_receiver <- as.factor(train_clean$set_receiver)
```

Distribution
```{r}
png(
file.path("_assets",
"set_rec2_dist.png"),
width=1080,
height=800,
pointsize=30 )

par(mar=c(4, 4, 4, 1))
data.frame(table(train_clean$set_receiver)) |>
    ggplot(aes(x = Var1, y = Freq)) +
    geom_bar(stat = "identity") +
    labs(title = "Grouped Set Receiver Distribution",
         x = "Set Receiver",
         y = "Frequency") + 
    theme_minimal()
dev.off()
```

### Kill
```{r}
png(
file.path("_assets",
"kill_dist.png"),
width=1080,
height=800,
pointsize=30 )

par(mar=c(4, 4, 4, 1))
data.frame(table(train_clean$kill)) |>
    ggplot(aes(x = Var1, y = Freq)) +
    geom_bar(stat = "identity") +
    labs(title = "Attack Success Distribution",
         x = "Attack Success",
         y = "Frequency") + 
    theme_minimal()
dev.off()
```

# Set Locaton Modeling

## Full Model

Model
```{r}
full_model <- multinom(set_loc ~ team1_diff + tot_point + set + game + team + level + game_sex + insys + back_setter,
                      data = train_clean)
summary(full_model, Wald.ratios = TRUE)
```

Z-scores and p-values
```{r}
z <- summary(full_model)$coefficients/summary(full_model)$standard.errors
p <- (1-pnorm(abs(z), 0, 1))*2
```

Prediction Accuracy
```{r}
p <- predict(full_model, train_clean)
tab <- table(p, train_clean$set_loc)
tab

(388 + 61 + 82 + 49)/1464
```

### Counts Validation
```{r}
#| eval: false
ftable(tab1 <- xtabs(~set_loc + team1_diff + tot_point + set + game + team + level + set_sex + game_sex + net_height + insys + back_setter,
                     data = train_clean))
```


## Random Forest Variable Selection
```{r}
rf_model <- randomForest(set_loc ~ team1_diff + tot_point + set + game + team + level + game_sex + insys + back_setter,
                      data = train_clean,
                 ntree = 1000,
                 mtry = 4,
                 importance = TRUE)
print(rf_model)
varImpPlot(rf_model)
```

```{r}
png(
file.path("_assets",
"rf_varImp_loc.png"),
width=1080,
height=800,
pointsize=30 )

par(mar=c(4, 4, 4, 1))
varImpPlot(rf_model)
dev.off()
```


## Reduced Model 


Model
```{r}
red_model <- multinom(set_loc ~ team1_diff + tot_point + set + game + insys + back_setter,
                      data = train_clean)
summary(red_model, Wald.ratios = TRUE)
```

Z-scores and p-values
```{r}
z <- summary(red_model)$coefficients/summary(red_model)$standard.errors
p <- (1-pnorm(abs(z), 0, 1))*2
p < 0.05
```

Prediction Accuracy
```{r}
p <- predict(red_model, train_clean)
tab <- table(p, train_clean$set_loc)
tab

(384 + 53 + 81 + 56)/1464
```

### Count Validation
```{r}
tab <- ftable(tab1 <- xtabs(~set_loc + level + insys + back_setter,
                            data = train_clean))

xtb <- as.data.frame(tab)

xttbl <-
xtable(xtb,
caption="Set location counts by reduced predictors",
label="tab:red_pred_setCount")

xxx <- print(xttbl, include.rownames=FALSE, table.placement="H") ### need LaTeX pkg float

writeLines( xxx, file.path("_assets", "red_pred_setCount.tex") )
```
still some empty cells look at just insys and back_setter


## 2 predictor model

Model
```{r}
small_model <- multinom(set_loc ~ insys + back_setter,
                      data = train_clean)
summary(small_model, Wald.ratios = TRUE)
```


Prediction Accuracy
```{r}
p <- predict(small_model, train_clean)
tab <- table(p, train_clean$set_loc)
tab

(538 + 0 + 0 + 0)/1464
```

## Return to reduced model 

4v5
```{r}
eq4v5 <- data.frame(
log_odds = summary(red_model)$coefficients[1,],
odds_ratio = exp(summary(red_model)$coefficients[1,])
) |>
  mutate(z = log_odds/summary(red_model)$standard.errors[1,]) |>
  mutate(p = (1-pnorm(abs(z), 0, 1))*2)

xttbl <-
xtable(eq4v5,
caption="Logit Model Coefficients: set location 4 vs. 5",
label="tab:eq4v5")

xxx <- print(xttbl, table.placement="H") ### need LaTeX pkg float

writeLines( xxx, file.path("_assets", "eq4v5.tex") )
```

4vbackRow
```{r}
eq4vbackRow <- data.frame(
log_odds = summary(red_model)$coefficients[2,],
odds_ratio = exp(summary(red_model)$coefficients[2,])
) |>
  mutate(z = log_odds/summary(red_model)$standard.errors[2,]) |>
  mutate(p = (1-pnorm(abs(z), 0, 1))*2)

xttbl <-
xtable(eq4vbackRow,
caption="Logit Model Coefficients: set location 4 vs. backRow",
label="tab:eq4vbackRow")

xxx <- print(xttbl, table.placement="H") ### need LaTeX pkg float

writeLines( xxx, file.path("_assets", "eq4vbackRow.tex") )
```


4vmidCourt
```{r}
eq4vmidCourt <- data.frame(
log_odds = summary(red_model)$coefficients[3,],
odds_ratio = exp(summary(red_model)$coefficients[3,])
) |>
  mutate(z = log_odds/summary(red_model)$standard.errors[3,]) |>
  mutate(p = (1-pnorm(abs(z), 0, 1))*2)

xttbl <-
xtable(eq4vmidCourt,
caption="Logit Model Coefficients: set location 4 vs. midCourt",
label="tab:eq4vmidCourt")

xxx <- print(xttbl, table.placement="H") ### need LaTeX pkg float

writeLines( xxx, file.path("_assets", "eq4vmidCourt.tex") )
```


# Set Receiver Modeling

## Random Forest Variable Selection 
```{r}
rf_model <- randomForest(set_receiver ~ game + team + level + game_sex,
                      data = train_clean,
                 ntree = 1000,
                 mtry = 4,
                 importance = TRUE)
print(rf_model)
varImpPlot(rf_model)
```

Plot
```{r}
png(
file.path("_assets",
"rf_varImp_receiver.png"),
width=1080,
height=800,
pointsize=30 )

par(mar=c(4, 4, 4, 1))
varImpPlot(rf_model)
dev.off()
```

## Model 1

Model
```{r}
red_model <- multinom(set_receiver ~ team1_diff + tot_point + set + team + insys + back_setter,
                      data = train_clean)
summary(red_model, Wald.ratios = TRUE)
```

Z-scores and p-values
```{r}
z <- summary(red_model)$coefficients/summary(red_model)$standard.errors
p <- (1-pnorm(abs(z), 0, 1))*2
p < 0.05
```

Prediction Accuracy
```{r}
p <- predict(red_model, train_clean)
tab <- table(p, train_clean$set_receiver)
tab

(39 + 183 + 7 + 154 + 124)/1464
```
34.63% accuracy


### Count Validation
```{r}
tab <- ftable(tab1 <- xtabs(~set_receiver + team + insys + back_setter,
                            data = train_clean))

xtb <- as.data.frame(tab)

xttbl <-
xtable(xtb,
caption="Set Receiver counts by reduced predictors",
label="tab:red_pred_recCount")

xxx <- print(xttbl, include.rownames=FALSE, table.placement="H") ### need LaTeX pkg float

writeLines( xxx, file.path("_assets", "red_pred_recCount.tex") )
```

Team counts
```{r}
tab <- ftable(tab1 <- xtabs(~set_receiver + team,
                            data = train_clean))

xtb <- as.data.frame(tab)

xttbl <-
xtable(xtb,
caption="Set Receiver counts by team",
label="tab:team_recCount")

xxx <- print(xttbl, include.rownames=FALSE, table.placement="H") ### need LaTeX pkg float

writeLines( xxx, file.path("_assets", "team_recCount.tex") )
```

## No team predictor

Model
```{r}
noTeam_model <- multinom(set_receiver ~ team1_diff + tot_point + set + insys + back_setter,
                      data = train_clean)
summary(noTeam_model, Wald.ratios = TRUE)
```

Z-scores and p-values
```{r}
z <- summary(noTeam_model)$coefficients/summary(noTeam_model)$standard.errors
p <- (1-pnorm(abs(z), 0, 1))*2
p < 0.05
```

Prediction Accuracy
```{r}
p <- predict(noTeam_model, train_clean)
tab <- table(p, train_clean$set_receiver)
tab

(0 + 129 + 0 + 211 + 85)/1464
```
29.03% accuracy

## 2 predictor model

Model
```{r}
small_model <- multinom(set_receiver ~ insys + back_setter,
                      data = train_clean)
summary(small_model, Wald.ratios = TRUE)
```


Prediction Accuracy
```{r}
p <- predict(small_model, train_clean)
tab <- table(p, train_clean$set_receiver)
tab

(0 + 129 + 0 + 204 + 93)/1464
```

# Kill

## Full model
```{r}
full_model <- glm(kill ~ team1_diff + tot_point + set + game + set_loc + num_block + set_receiver + team + level + game_sex + insys + back_setter,
                   data = train_clean,
                   family = "binomial")
summary(full_model)
```

## Forward/Backward Propagation
```{r}
null_model <- glm(kill~1,
                  data = train_clean,
                  family = "binomial")
forward_model <- step(null_model,
                   scope = formula(full_model),
                   direction = "forward",
                   data = train_clean)
backward_model <- step(full_model,
                       direction = "backward",
                       data = train_clean)
```
Forward: insys + game_sex + set_receiver + num_block  -    AIC: 1854.56 
Backward: game + num_block + set_receiver + insys     -    AIC: 1861.37




Summary
```{r}
summary(forward_model)
```

$R^2$
```{r}
1 - (1838.6/1952.7)
```

Accuracy
```{r}
p <- predict(forward_model, train_clean, type = "response")
pred <- ifelse(p > 0.5, 1, 0)
sum(pred)/1464
```
15.399% accuracy

###Goodness of fit
```{r}
sum(residuals(forward_model,type="pearson")^2)
qchisq(0.95, 1456)
```

### Model table
```{r}
forward_model_ests <- data.frame(summary(forward_model)$coefficients) |>
  mutate(sig = `Pr...z..` < 0.05,
         odds = exp(Estimate)) |>
  select(Estimate, odds, everything())
names(forward_model_ests) <- c("log_odds", "odds", "Std. Error", "Z-stat", "p-value", "Sig")

xtb <- forward_model_ests

xttbl <-
xtable(xtb,
caption="Hit Result Model Coefficients",
label="tab:eqKill")

xxx <- print(xttbl, table.placement="H") ### need LaTeX pkg float

writeLines( xxx, file.path("_assets", "eqKill.tex") )
```

### Assumptions
```{r}
library(car)
library(alr4)

png(
file.path("_assets",
"kill_resPlot.png"),
width=1080,
height=800,
pointsize=30 )

par(mar=c(4, 4, 4, 1))
suppressWarnings(residualPlots(forward_model))
mtext("Attack Success Residual Plots", side = 3, line = 3)
dev.off()


png(
file.path("_assets",
"kill_inflPlot.png"),
width=1100,
height=1000,
pointsize=30 )

par(mar=c(4, 4, 4, 1))
suppressWarnings(influenceIndexPlot(forward_model, vars=c("Cook","hat"), id.n=3))
dev.off()
```

# Model Validation
```{r}
test_clean <- test |>
  select(-c(points1, points2, team2_diff)) |>
  mutate(insys = ifelse(pass_loc %in% c(2,3,4,7) & pass_quality %in% c(4,5),
                        TRUE,
                        FALSE),
         back_setter = ifelse(rotation %in% c(1, 2, 3),
                              TRUE,
                              FALSE)) |>
  select(-c(pass_loc, pass_quality, receiver_loc, rotation)) |>  
  mutate(set_loc = case_when(
    set_loc %in% c("back1", "1", "2", "back2", "3", "dump") ~ "midCourt",
    set_loc == "back3" ~ "5",
    set_loc %in% c("A", "B", "C", "D", "pipe") ~ "backRow",
    TRUE ~ set_loc
  )) |>
  mutate(set_loc = as.factor(set_loc)) |>
  mutate(set_receiver = case_when(
    set_receiver %in% c("ds", "libero", "setter1", "setter2", "opposite2", "outside3") ~ "other",
    set_receiver %in% c("middle1", "middle2", "middle3") ~ "middle",
    set_receiver == "outside" ~ "outside2",
    TRUE ~ set_receiver
  ))
  
  
  
  
# make a 4 the reference category(default set)
test_clean$set_loc <- relevel(test_clean$set_loc, ref = "4")  
```



## Set Location
```{r}
red_model <- multinom(set_loc ~ team1_diff + tot_point + set + level + insys + back_setter,
                      data = train_clean)
```

Test set accuracy
```{r}
p <- predict(red_model, test_clean)
tab <- table(p, test_clean$set_loc)
tab

(157 + 15 + 0 + 10)/488
```


## Kill
```{r}
p <- predict(forward_model, test_clean, type = "response")
pred <- ifelse(p > 0.5, 1, 0)
sum(pred)/518
```
13.707% accuracy


















